The Practical Application of Computerised Multistage Tests

### 2021-03-04 article.Rmd
Initial edits from Sz corrections

---
title: "The Practical Application of Computerised Multistage Tests"
author:
- name: James Page
  affiliation: SP Educational Consulting
  affiliation_url: https://www.speduconsulting.co.uk/
date: "`r Sys.Date()`"
output:
  pdf_document:
    highlight: pygments
    latex_engine: lualatex
  html_document:
    df_print: paged
version: 1
description: A practical guide to the design and development of computerised multistage
  tests.
bibliography: references.bib
---

@@

The advantages of splitting tests up into units of items called *modules* and directing candidates to these modules based on their latent ability estimates, provides a more efficient test delivery than a corresponding linear test, whilst also offering an easier method of assembly, and greater protection against test copying.

I'll assume some familiarity with Item Response Theory (IRT) - and the Rasch model. Some knowledge of programming with R would be useful to understand the examples given. For background information and primers on these subjects I provide references.

My main focus is the development of tests for language learners, specifically English language learners. The theory and data processing involved can  be transferred to other domains that require measurement of latent ability. The IRT model that I focus on is the 1PL (or Rasch) model. For readers interested in extending analysis to more complex models, see @magis2017computerized.

Developing a new test is a complex, multi-disciplinary task, involving specialist item writers, psychometricians, software engineers, validation experts and project managers. It is essential, before beginning any of this work to have a clear vision of what you are seeking to test, and the types of candidates that you will be targeting along with their needs. The main tasks when developing a new test will include:

* Blueprint, test specification and test design
* Simulation
* Item writing
* Item review and rating
* Pilot/Live trial
* Rasch modelling and item banking
* Cut scores and scale scores
* Live test

Once the live test has been deployed, continual monitoring and adjustment of the test will be essential to ensure that it remains secure and delivers on the targetted goals over the long term.

@@

In addition the test delivery system may be restricted in its ability to access the item bank to derive item difficulties. In this case, mapping functions to derive ability estimates and scale scores as well as provision of the cut scores for module routing must be provided in advance of test administration for each panel. We demonstrate methods that can be used to provide these values.

@@

The underlying theoretical basis for MST is the branch of psychometrics known as Item Response Theory, or IRT. The theory proposes that there are *latent traits* (for example language ability) that can be measured using a set of items delivered to the test candidate. These test items are typically scored as correct (1) or incorrect (0) - and it is these dichotomous test types that we focus on here. It is also possible to offer polytomous test types (for example based on Likert scales), but we won't cover those in this article.

The response to the test items can then be used to estimate the candidate's latent ability. This can be done by first fitting an IRT model that can then be used to make predictions about latent ability based on the item response. Once the latent ability estimate has been calculated, this can then be converted into a scale score, an ability summary, or both. The rest of this document provides practical examples of designing a multistage test, fitting an IRT model and calculating scores.

@@

For every item, the probability that the item will be answered correctly can be visualised by plotting an Item Characteristic Curves (ICCs), which uses the above logistic function.

![Item Characteristic Curves](images/icc.png)

@@

The slope of the curve indicates how discriminating the item is. An item that is very discriminating will have a steep curve; one that is less discriminating will have a flatter curve. Highly discriminating items have more information. Ideally you want to have many discriminating items to be able to provide a lower error in your ability estimate. In general, the more items you are able to trial with the more information (and less error) you will have.

@@

An item that all candidates always get right because it is too easy, or they always get wrong because it's too hard has very little informative value for a test. It's telling you very little about the ability levels of your candidates. From this it follows that an item where half the candidates answer correct and the other half answer wrong has the most informational value; this is the point where $\theta$ and $b$ are equal. This relationship can be expressed by the Item Information Function (IIF). In the 1PL model the maximum value of the IIF is 0.25 (it can be different in the other models due to the influence of the $a$ parameter).

@@

As mentioned, greater information implies a smaller error of measurement. When designing a test it is important to deliver items with a range of difficulties that match the expected ability ranges so that you can accurately measure the your target candidates. In the above diagram you can see that item 587 (blue) is good at measuring candidates around ability level 0.95. Item 583 is better at measuring candidates with a lower ability level.

@@

The log odds, or *logit* of a correct response can be expressed (in the Rasch model) as $\theta - b$; so if a candidate has ability $\theta$ which happens to be equal to the difficulty of our item $b_{583}$, then the odds are even (1:1) that they will answer it correctly. The logit value here would be 0 (remember this is also the point of maximum information for the item). The logit function is the inverse of the logistic function described earlier:

@@

The item response function is a logistic function with a characteristic sigmoid (or 'S' shape), and can be plotted for each item. Throughout the rest of this article we'll consider an example item which will follow a *lifecycle* from when it is first conceived, developed, rated and trialled, banked in the item bank, then delivered as part of one or more test administrations, and finally retired from use.

Of course, this item, let's call it 'Item583', will be one of many that will, over time, be added into an item bank. The item will have a number of properties and metadata that describes it. For example the name of the item writer who created it, a set of media (audio, video or image files) that is associated with it, and importantly for our purposes, the difficulty coefficient $b$. The rest of this article follows this lifecycle, in particular how it relates to the various IRT methodologies for design and validation of multistage tests.

@@

When designing a multistage test, its important to ensure that items are written to the target ability that you want to measure.

Modules need to be constructed in such a way that they target a specified ability range. We'll see later on how a Module Information Function can be used to help with module routing decisions. Also, routing needs to be designed with overlapping items between adjacent modules.

@@

IRT is used to inform calibration of the test items after they've been developed by subject matter experts, as well as construction of the modules and routing between them. Scoring of the test can be derived from the $\theta$ estimates. Typically the initial module (known as the routing module) will have an overall difficulty of 0 (ie. a mix of easy and hard items), with subsequent routing directing candidates to easier or more difficult modules in order to 'hone in' on an accurate $\theta$ estimate at the end of the test.

A typical modular design will look like this:

@@

This is an example of a 1-2-4 design, but others are possible. Important questions to answer during test design are what module topology to use and what should be the overall difficulty levels for each module (represented as the mean item difficulty, $\bar{b}$).

@@

A benefit of using modules is that items can be placed into modules and then delivered as parallel test forms called *panels*. Having multiple panels helps to reduce item exposure rate as previously seen panels can be precluded from selection for future test administrations. It also makes it easier to assemble tests. It is possible to place items in more than one panel, but its important to ensure that parallel panels are of equivalent difficulty.

@@

When constructed in this way each module can be considered as a module instance (for example Module A:3 being the instance of module A in panel 3),

# Simulation

In the pre-trial stage, it is advisable to run simulations to help us answer these questions about test design. It's always less expensive to make modifications to a design early on in the process, before significant investment in item writing and trialling has been made. The ability to simulate a test at the design phase can be used to understand the constraints of the design at an early stage and make changes in a low-cost, low-risk way. In this section we'll show how the `mstR` R package can be used to run simulations for a MST design. First we'll discuss the data sources that we'll be using.

@@

I mentioned earlier that multistage test designs comprise a number of modules arranged into stages. The question then becomes, how many stages do I need, and how many modules in each stage do I want? We saw earlier an example of a simple 1-2-4 design.

Normally the so-called routing module in stage 1 (Module A) would contain a mix of items with a mean difficulty of 0 (in other words 'medium' difficulty). Candidates that are placed in the lower ability range at the end of this module will be routed to the 'easy' module in stage 2 (Module B), and those in the upper ability range to the 'hard' module (Module C).

@@

Assuming the 1-2-4 pattern shown above, a simulation for the test design can be created from following elements:

@@

Average difficulty levels for each module are shown below (where $\bar{b}$ is the mean difficulty of the module, and a standard deviation of 1 is assumed)

![MST Design](images/module-items.png)

@@

Here we've decided to use the 1PL (or Rasch) model, so we need to supply `bPrior` difficulty distributions, and convert the data frame into a matrix. The `bPrior` parameter is constructed using a normal distribution with a given $\bar{b}$ and standard deviation of 1.
We'll end up with a matrix of 68 items, each with randomised difficulty parameters. A test will comprise 30 items (12 + 8 + 10).

@@

Selecting items into the modules is done by creating another matrix with 68 rows (items) and 7 columns (modules); a cell containing a 1 indicates membership of an item in a module.

```{r module items, echo=TRUE}
modules <- matrix(0, 68, 7)
modules[1:12, 1] <- 1
modules[13:20, 2] <- 1
modules[21:28, 3] <- 1
modules[29:38, 4] <- 1
modules[39:48, 5] <- 1
modules[49:58, 6] <- 1
modules[59:68, 7] <- 1
```

@@

The final part of the design activity is construction of the routing paths and cut-offs used for next module selection. This is done by creating an $R \times R$ *transition matrix* which shows the valid paths to modules in the next stage (where $R$ is the number of modules). A value of 1 in the `routing` matrix in the code sample below indicates that a path exists from module $i$ to module $j$. To use the example below the rule `routing[1, 2:3] <- 1` means module 1 routes to modules 2 and 3.

In addition to the routing paths, a set of cut-off criteria must be given. These are expressed in terms of $\hat{theta}$ using the `cutoff` matrix. To use an example here, the rule `cutoff[2,] <- c(-Inf, 0.0)` means that candidates with ability estimates from minus infinity to zero will be routed to module 2. The rule `cutoff[3,] <- c(0.0, Inf)` means that candidates with ability estimates from zero to plus infinity will be routed to module 3.

```{r module routing, echo=TRUE}
routing <- matrix(0, 7, 7)
routing[1, 2:3] <- 1
routing[2, 4:6] <- 1
routing[3, 5:7] <- 1

cutoff <- matrix(NA, 7, 2)
cutoff[2,] <- c(-Inf, 0.0)
cutoff[3,] <- c(0.0, Inf)
cutoff[4,] <- c(-Inf, -1.0)
cutoff[5,] <- c(-1.0, 0.0)
cutoff[6,] <- c(0.0, 1.0)
cutoff[7,] <- c(1.0, Inf)

@@

As can be seen, modules 4-7 (Module D to Module G in the design) are final stage modules, so do not route onward.

@@

Here we use a $\theta$ of `r theta` as an example. Note that this doesn't really simulate a true MST test, as it ignores the routing rules and simply returns a response for all items in the simulated item bank. We can however use the `randomMST()` function to perform a more realistic simulation. The `randomMST()` function will generate a response pattern for a given simulated item bank, an MST structure, and a true ability level ($\theta$). We'll use Bayesian model estimation (`"BM"`) for ability estimation (the default ability estimation method for the `mstR` library). The maximum Fisher information criterion (`"MFI"`) for is used for module selection; again this is the default. The Fisher information is the measure of the amount of information that item $X$ carries about the unknown ability parameter $\theta$.

@@

For this simuluted candidate with a true $\theta$ = `r dfMSTResponse$trueTheta`, the estimated ability $\hat{\theta}$ = `r dfMSTResponse$thFinal`. Note that the output from the `randomMST` function uses the module and item indices (rather than names), so you will want to map these back to a name for ease of interpretation.

The Module Information Function is used to determine the 'best' (that is the most informative) module to route to in the next stage. In this example the selected modules are `r dfMSTResponse$selected.modules`. A plot can be made as follows, showing the highlighted route through the multi-stage test:

>>>>

---
title: "The Practical Application of Computerised Multistage Tests"
description: A practical guide to the design and development of computerised multistage
  tests.
author:
- name: James Page
  affiliation: SP Educational Consulting
  affiliation_url: https://www.speduconsulting.co.uk/
date: "`r Sys.Date()`"
version: 1
output:
  pdf_document:
    highlight: pygments
    latex_engine: lualatex
    toc: true
    toc_depth: 2
    fig_caption: true
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
  word_document:
    toc: true
    toc_depth: 2
bibliography: references.bib
---

@@

This article is split into 3 main parts:
1.	The design of MSTs
2.	The development of MSTs
3.	The validation of MSTs
  a.	Test construction
  b.	Trialling
  c.	Scoring and routing

In this article we’ll consider an example *item* which will follow a lifecycle from when it is first conceived, developed, rated and trialled, banked in the item bank, delivered as part of one or more test administrations, and finally retired from use. We will also consider a hypothetical *candidate* and follow their *journey* from being entered to take the test, being routed between the different stages of the test, and receiving a final score and report on their results.

# The Design of MSTs

The advantages of splitting tests up into units of items called *modules* and directing candidates to these modules based on their latent ability estimates provides a more efficient test delivery than a corresponding linear test, whilst also offering an easier method of assembly, and greater protection against test copying.

We'll assume some familiarity with Item Response Theory (IRT) - and the Rasch model. Some knowledge of programming with R would be useful to understand the examples given. For background information and primers on these subjects we provide references.

The main focus is the development of tests for language learners, specifically English language learners. The theory and data processing involved can  be transferred to other domains that require measurement of latent ability. The IRT model that we focus on is the 1PL (or Rasch) model. For readers interested in extending analysis to more complex models, see @magis2017computerized.

Developing a new test is a complex, multi-disciplinary task, involving content experts, specialist item writers, psychometricians, software engineers, validation experts and project managers. It is essential, before beginning any of this work to have a clear vision of what you are seeking to test, and the types of candidates that you will be targeting along with their needs. The main tasks when developing a new test will include:

* Blueprint, test specification and test design
* Simulation
* Item writing
* Item review and rating
* Pilot/Live trial
* Rasch modelling and item banking
* Establishing cut scores and scale scores
* Live test

Once the live test has been launched, continual monitoring and adjustment of the test will be essential to ensure that it remains secure and delivers on the targeted goals over the long term.

@@

In addition, the test delivery system may be restricted in its ability to access the item bank to derive item difficulties. In this case, mapping functions to derive ability estimates and scale scores as well as provision of the cut scores for module routing must be provided in advance of test administration for each panel. We demonstrate methods that can be used to provide these values.

@@

The underlying theoretical basis for MST is the branch of psychometrics known as Item Response Theory, or IRT. The theory proposes that there are *latent traits* (for example language ability) that can be measured using a set of items delivered to the test candidate. These test items are typically scored as correct (1) or incorrect (0) - and it is these dichotomous item types that we focus on here. It is also possible to offer polytomous item types (for example based on Likert scales, or partial credit items), but we won't cover those in this article.

The responses to the test items can then be used to estimate the candidate's latent ability. This can be done by first fitting an IRT model to a response dataset collected during the trial phase. This model can then be used to make predictions about latent ability based on item responses once the test goes live. The latent ability estimates can then be converted into a scale score, an ability summary, or both. The rest of this document provides practical examples of designing a multistage test, fitting an IRT model and calculating scores.

@@

For every item, the probability that the item will be answered correctly can be visualised by plotting an Item Characteristic Curve (ICC), which uses the above logistic function. Figure 1 plots the ICCs for three items: Item579 (red), Item583 (green) and Item587 (blue).

![Item Characteristic Curves \label{Figure 1}](images/icc.png)

@@

The slope of the curve indicates how discriminating the item is. An item that is very discriminating will have a steep curve; one that is less discriminating will have a flatter curve. Highly discriminating items have more measurement information. Ideally you want to have many discriminating items to be able to provide a lower error in your ability estimate. In general, the more items you are able to trial with, the more information (and less error) you will have.

@@

An item that all candidates always get right because it is too easy, or they always get wrong because it's too hard has very little informational value for a test. It's telling you very little about the ability levels of your candidates. From this it follows that an item where half the candidates answer correct and the other half answer wrong has the most informational value; this is the point where $\theta$ and $b$ are equal, so the item is perfectly targeted at the ability of these candidates. This relationship can be expressed by the Item Information Function (IIF). In the 1PL model the maximum value of the IIF is 0.25 (it can be different in the other models due to the influence of the discrimination parameter $a$).  Figure 2 displays the IIFs for the same three items as before.

@@

As mentioned, greater information implies a smaller error of measurement. When designing a test it is important to deliver items with a range of difficulties that match the expected ability ranges of the candidates so that you can accurately measure the your target candidates. In the above diagram you can see that Item 587 (blue) is good at measuring candidates around ability level 0.95. Item (green) 583 is better at measuring candidates with a lower ability level.

@@

The log odds, or *logit* of a correct response can be expressed (in the Rasch model) as $\theta - b$; so if a candidate has ability $\theta$ which happens to be equal to the difficulty of our item $b_{583}$, then the odds are even (1:1) that they will answer it correctly (in other words a 50% chance of getting it right, and a 50% chance of getting it wrong). The logit value here would be 0 (remember this is also the point of maximum information for the item). The logit function is the inverse of the logistic function described earlier:

@@

The item response function is a logistic function with a characteristic sigmoid (or 'S' shape), and can be plotted for each item. Throughout the rest of this article we'll consider an example item, **Item583**, which will follow a *lifecycle* from when it is first conceived, developed, rated and trialled, banked in the item bank, then delivered as part of one or more test administrations, and finally retired from use.

Of course, this item is just one of many that will, over time, be added into an item bank. The item will have a number of properties and metadata that describes it. For example the name of the item writer who created it, the skill it is targeting, the sub-skills required, a set of media (audio, video or image files) that is associated with it and which serve as stimuli, the instructions, the prompt question, and importantly for our purposes, the difficulty coefficient $b$. The rest of this article follows this life cycle, in particular how it relates to the various IRT methodologies for design and validation of multistage tests.

@@

When designing any test, it is important to ensure that items are written to the target ability range that you want to measure. Subject Matter Experts (SMEs) will need to be involved in the initial test design to make decisions about the length of the test and the components (skills as R,L,S,W, or any combination of these) for the target candidature.

In a multistage test, modules need to be constructed in such a way that they target a specified ability range. We'll see later on how a Module Information Function can be used to help with module routing decisions. Also, routing needs to be designed with overlapping items between adjacent modules.

@@

IRT is used to inform calibration of the test items after they've been developed by item writers, as well as construction of the modules and routing between them. Scoring of the test can be derived from the ability estimates. Typically the initial module (known as the routing module) will have an overall difficulty of 0 (ie. a mix of easy and hard items), with subsequent routing directing candidates to easier or more difficult modules in order to 'hone in' on an accurate $\hat{theta}$ (ability estimate) at the end of the test.

A typical modular design for a MST will look like this (Figure 3):

@@

This is an example of a 1-2-4 design, but others are possible. Important questions to answer during test design are what module topology to use (the number of stages and modules per stage), and what should be the overall difficulty levels for each module (represented as the mean item difficulty, $\bar{b}$).

As mentioned earlier, multistage test designs comprise a number of modules arranged into stages. The question then becomes, how many stages do we need, and how many modules in each stage do we want? We saw earlier an example of a simple 1-2-4 design.

Normally the so-called routing module in stage 1 (Module A) would contain a mix of items with a mean difficulty of 0 (in other words ‘medium’ difficulty). Candidates that are placed in the lower ability range at the end of this module will be routed to the ‘easy’ module in stage 2 (Module B), and those in the upper ability range to the ‘hard’ module (Module C).

@@

A benefit of using modules is that items can be placed into modules and then delivered as parallel test forms called *panels*. Having multiple panels helps to reduce item exposure rate as previously seen panels can be precluded from selection for future test administrations. It also makes it easier to assemble tests. It is possible to place items in more than one panel, but its important to ensure that parallel panels are of equivalent difficulty, see Figure 4.

@@

When constructed in this way each module can be considered as a module instance (for example Module D:3 being the instance of module D in panel 3), Once the test topology has been decided, the first panel can be developed a prototype.

# The Development of MSTs

In this part we describe the steps needed in the development of MSTs. Here we turn our attention to initial validation to confirm the soundness of the test design that was conceived at the first stage of test development described in Part 1 of this article.

## Simulation

In the pre-trial stage, it is advisable to run simulations to help us answer questions raised in Part 1 of this article, and inform final decisions about test design. It's always less expensive to make modifications to a design early on in the process, before significant investment in item writing and trialling has been made. The ability to simulate a test at the design phase can be used to understand the constraints of the design at an early stage and make changes in a low-cost, low-risk way. In this section we'll show how the `mstR` R package can be used to run simulations for a MST design. First we'll discuss the data sources that we'll be using.

@@

@@

Assuming the 1-2-4 pattern shown above, with a total of 30 items (12 + 8 + 10), a simulation for the test design can be created from following elements:

@@

Average difficulty levels for each module are shown in Figure 5 below (where $\bar{b}$ is the mean difficulty of the module, and a standard deviation of 1 is assumed)

![MST Design with average difficulty levels](images/module-items.png)

@@

Here we've decided to use the 1PL (or Rasch) model, so we need to supply `bPrior` difficulty distributions, and convert the data frame into a matrix. The `bPrior` parameter is constructed using a normal distribution with a given $\bar{b}$ and standard deviation of 1.
We'll end up with a matrix of 68 items, each with randomised difficulty parameters.

@@

Selecting items into the modules is done by creating another matrix with 68 rows (items) and 7 columns (modules); a cell containing a 1 indicates membership of an item in a module. Because we're using an R matrix to store this information we must use the index of the module (module A == 1, module B == 2, and so on); the code snippet below uses a helper function called `module()` to map from names to indices.

```{r module items, echo=TRUE}
# create a function to map module names to indices
moduleIndices <- seq(1:7)
names(moduleIndices) <- c("A", "B", "C", "D", "E", "F", "G")
module <- function(n) { unname(moduleIndices[n]) }

modules <- matrix(0, 68, 7)
modules[1:12, module('A')] <- 1
modules[13:20, module('B')] <- 1
modules[21:28, module('C')] <- 1
modules[29:38, module('D')] <- 1
modules[39:48, module('E')] <- 1
modules[49:58, module('F')] <- 1
modules[59:68, module('G')] <- 1
```

@@

The final part of the design activity is construction of the routing paths and cut-offs used for next module selection. This is done by creating an $R \times R$ *transition matrix* which shows the valid paths to modules in the next stage (where $R$ is the number of modules). A value of 1 in the `routing` matrix in the code sample below indicates that a path exists from module $i$ to module $j$. To use the example below the rule `routing[module('A'), module('B'):module('C')] <- 1` means module A routes to modules B and C.

In addition to the routing paths, a set of cut-off criteria must be given. These are expressed in terms of $\hat{theta}$ using the `cutoff` matrix. To use an example here, the rule `cutoff[module('B'),] <- c(-Inf, 0.0)` means that candidates with ability estimates from minus infinity to zero will be routed to module B. The rule `cutoff[module('C'),] <- c(0.0, Inf)` means that candidates with ability estimates from zero to plus infinity will be routed to module C.

```{r module routing, echo=TRUE}
routing <- matrix(0, 7, 7)
routing[module('A'), module('B'):module('C')] <- 1
routing[module('B'), module('D'):module('F')] <- 1
routing[module('C'), module('E'):module('G')] <- 1

cutoff <- matrix(NA, 7, 2)
cutoff[module('B'),] <- c(-Inf, 0.0)
cutoff[module('C'),] <- c(0.0, Inf)
cutoff[module('D'),] <- c(-Inf, -1.0)
cutoff[module('E'),] <- c(-1.0, 0.0)
cutoff[module('F'),] <- c(0.0, 1.0)
cutoff[module('G'),] <- c(1.0, Inf)

@@

As can be seen, module D to module G are final stage modules, so do not route onward.

@@

Here we use a $\theta$ of `r theta` for our hypothetical candidate as an example. Note that this doesn't really simulate a true MST test, as it ignores the routing rules and simply returns a response for all items in the simulated item bank. We can however use the `randomMST()` function to perform a more realistic simulation. The `randomMST()` function will generate a response pattern for a given simulated item bank, an MST structure, and a true ability level ($\theta$). We'll use Bayesian model estimation (`"BM"`) for ability estimation (this is the default ability estimation method for the `mstR` library). The maximum Fisher information criterion (`"MFI"`) for is used for module selection; again this is the default. The Fisher information is the measure of the amount of information that item $X$ carries about the unknown ability parameter $\theta$.

@@

For this simulated candidate with a true $\theta$ = `r dfMSTResponse$trueTheta`, the estimated ability $\hat{\theta}$ = `r dfMSTResponse$thFinal`.

The Module Information Function is used to determine the 'best' (that is the most informative) module to route to in the next stage. In this case the modules that the candidate was routed to were `r names(moduleIndices[dfMSTResponse$selected.modules])`. A plot can be made as follows, showing the highlighted route through the multi-stage test. Note that the output from the `randomMST` function and the plot use the module and item indices (rather than their labels), so you may want to map these back to a name for ease of interpretation.
